---
title: "README"
author: "Josiah Parry"
date: "5/26/2017"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
my_access_token <- "ml7d-06dREg7ZNTrdG2CQ1wfoJ2m1FSP6ZRGIphEj2RRGtY_gpKiCfFFWpeMpTbC"
```

# geniusR

Welcome to `geniusR`, an R wrapper for the [Genius](https://genius.com/Genius-about-genius-annotated) API. The functions I created here were to serve my desires for acquiring song lyrics to be used in a tidy text analysis. Since song lyrics are lisenced, the **Genius** API cannot directly provide lyrics to any given song. Therefore, I utilized the search functions of their API to extract song url's and scrape their website for their lyrics. 

## Pre-requisites

Prior to being able to use `geniusR` you will first need to [create an account with Genius](https://docs.genius.com/) to get an API access token. 

## Using geniusR

Since I just developed these functions today (05.26.17), it isn't a real package yet. You can clone this repo and run each script to get the functions into you environment. Then, getting song lyrics is quite easy.

The main function you will be using to acquire song lyrics is `genius_lyrics()`.

## `genius_lyrics()`

First we will need to load our libraries and the functions to access the api. 
```{r, message=FALSE, warning=FALSE}
# Load libraries
library(tidyverse)
library(rvest)
library(stringr)

# Load all functions
sapply(file.path("R",list.files("R")), FUN = source)
```

The function should be pretty intuitive. Where the first argument is the artist name, second is the song, and the third is your access token. Below I access Kendrick Lamar's new hit from his album **DAMN.**. 

```{r, message=FALSE, warning=FALSE}
HUMBLE <- genius_lyrics(artist = "Kendrick Lamar", song = "HUMBLE", access_token = my_access_token)
```

```{r}
HUMBLE %>%
  mutate(song = "HUMBLE.", 
         album = "DAMN.",
         artist = "Kendrick Lamar")
```

This is my first ever attempt at accessing an API through R (or ever for that matter). As you can see above there are still many bugs. Issues and pull requests are welcome!