# Overview

This package was created to provide an easy method to access lyrics as text data using the website [Genius](genius.com). This package contains functions for getting:
  - Song lyrics
  - Album tracklists
  - Lyrics for an entire album
  
## Installation  
This package must be installed from GitHub.

```{r, eval = F}
devtools::install_github("josiahparry/geniusR")
```

## Song Lyrics

The main function you will be using to acquire song lyrics is `genius_lyrics()`. Supply an artist and a song title as character strings, and voila.

```{r}
library(geniusR)
genius_lyrics(artist = "Margaret Glaspy", song = "Memory Street")
```

This function relies on two other functions used internally, `gen_song_url()`, and `genius_url()`. The former function creates a character string to be supplied to `genius_url()` which scrapes lyrics from [Genius](genius.com).

### `genius_url()`

This function is intended to take the output from `gen_song_url()` (which is to be understood as _"generate song url"_). `gen_song_url()` takes two arguments `artist` and `song`. 

Example:
```{r}
song_url <- gen_song_url(artist = "Now, Now,", song = "Dead Oaks")
song_url

#genius_url(song_url)
```

## Albums

There are two functions to be used in accessing information about albums: `genius_album()` and `genius_tracklist()`. 

### `genius_tracklist()`

This function, given an `artist` and an `album` will return a barebones `tibble` with title and track number. 

```{r}
genius_tracklist(artist = "Basement", album = "Colourmeinkindness") 
```

### `genius_album()` 

Often times when doing an analysis of music, it becomes handy to have a whole album. This function will retrieve all lyrics for each track in an album in a tidy(ish) format. `genius_album()` returns a tibble with `track number`, `title`, and `lyrics` as a **nested** `tibble` column. 

`genius_album()` has a `nested` argument. Provide `nested = FALSE` to return a `tibble` with expanded lyrics where each row is one line of one song. 

```{r}
genius_album(artist = "From Indian Lakes", album = "Absent Sounds", nested = TRUE)
```

Internally, this works similarly to `genius_lyrics()`. It first utilizes the `gen_album_url()` function to create a url to the Genius page with the tracklist. It then utilizes `genius_tracklist()` and feeds each track title to `genius_lyrics()` to retrieve each individual song via `purrr::map()`. 

### Notes:

As this is my first _"package"_ there will be many issues. Please submit an issue and I will do my best to attend to it. 

There are already issues of which I am present (the lack of error handling). If you would like to take those on, please go ahead and make a pull request. Please contact me on [Twitter](twitter.com/josiahparry).
