---
title: "geniusR Basics"
author: "Josiah Parry"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# geniusR

geniusR enables qucik and easy download of song lyrics. The intent behind the package is to be able to perform text based analyses on songs in a tidy[text] format. 

This package was inspired by the release of Kendrick Lamar's most recent album, **DAMN.**. As most programmers do, I spent way too long to simplify a task, that being accessing song lyrics. Genius (formerly Rap Genius) is the most widly accessible platform for lyrics. 

The functions in this package enable easy access of individual song lyrics, album tracklists, and lyrics to whole albums. 

## Individual songs `genius_lyrics()`

Getting lyrics to a single song is pretty easy. Let's get in our **ELEMENT.** and checkout **DNA.**. But first, note that the `genius_lyrics()` function takes two arguments, `artist` and `song`. Be sure to spell the name of the artist and the song correctly, but don't worry about capitalization. 

First, let's set up our libraries / working environment. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load needed libraries
library(geniusR)
library(tidyverse)
library(tidytext)
```

`genius_lyrics()` returns only the barebones. Utilizing `dplyr` we can also create a new variable with the line number to help in future [tidytext](https://github.com/juliasilge/tidytext) analysis. This will be covered in a later vignette / post.  

```{r}
DNA <- genius_lyrics(artist = "Kendrick Lamar", song = "DNA.")
DNA %>% 
  mutate(line = row_number())

```

## Album Level Information

### Tracklists

There are two key functions to be utilized at the album level: `genius_tracklist()` and `genius_album()`.

I often only know an album name and none of the track titles. Often I only know the position in the tracklist. For this reason, I created a tool to provide an album tracklist. This function, `genius_tracklist()` takes the arguments `artist` and `album`. Simple enough, right? 

Let's get the tracklist for the original release of **DAMN.**. However, real Kendrick fans know that the album was intended to be listened to in chronological *and* reverse order—as is on the collector's release. 


```{r}
damn_tracks <- genius_tracklist(artist = "Kendrick Lamar", album = "DAMN.")

# Collector's reverse order
damn_tracks %>% 
  arrange(-track_n)
```

### Album Lyrics

If lyrics for a full album are what you desire, look no further than `genius_album()`. With a little help from [`purrr`](https://github.com/tidyverse/purrr) you can avoid the annoying iterations! `genius_album()` is designed for simple use. Provide it with only two argument—`artist` and `album`—and you'll have lyrics in no time! 

`genius_album()` returns a tibble with a column of nested data frames where each song's lyrics are contained. Nesting the lyrics was done to have a less cluttered data frame. You can unnest the lyrics by setting `nested = FALSE`. Alternatively, you can pipe it into an `unnest()` call. 


```{r}
DAMN <- genius_album(artist = "Kendrick Lamar", album = "DAMN.", nested = FALSE)

# Alternative method
#genius_album(artist = "Kendrick Lamar", album = "DAMN.") %>% 
#  unnest(lyrics)

head(DAMN)
```

Bam. Easy peasy. Now you have a sweet data frame ready for a tidy text analysis!


