---
title: "geniusR Basics"
author: "Josiah Parry"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# geniusR

geniusR enables quick and easy download of song lyrics. The intent behind the package is to be able to perform text based analyses on songs in a tidy[text] format. Song lyrics come from [Genius](https://genius.com) (formerly Rap Genius), the most widely accessible platform for lyrics. 

The functions in this package enable easy access of individual song lyrics, album tracklists, and lyrics to whole albums. 

## Individual songs `genius_lyrics()`

Getting lyrics to a single song is pretty easy. Let's get in our **ELEMENT.** and checkout **DNA.**. But first, note that the `genius_lyrics()` function takes two arguments, `artist` and `song`. Be sure to spell the name of the artist and the song correctly, but don't worry about capitalization. 

First, let's set up our working environment. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load needed libraries
library(geniusR)
library(dplyr)
```

`genius_lyrics()` returns only the barebones. Utilizing `dplyr` we can also create a new variable with the line number to help in future [tidytext](https://github.com/juliasilge/tidytext) analysis. This will be covered in a later vignette / post.  

```{r}
DNA <- genius_lyrics(artist = "Kendrick Lamar", song = "DNA.")
DNA

```

This function also enables you to get verse level information. This tends to me more popular in hip-hop where songs may have multiple feature artists (thanks to [\@natebarr64](https://github.com/natebarr64) for the help!). `genius_lyrics()` has an argument called `info`. It defaults to `"title"`. If you provide the argument `info = "features"` two new columns appear, `"verse"` and `"vocalist"`.

```{r}
genius_lyrics(artist = "Kendrick Lamar", song = "DNA.", info = "features")
```


## Albums

More often than not you will want to get the lyrics for an entire album. This is done easily with `genius_album()`. Just provide the `artist` and `album` name.

```{r}
DAMN <- genius_album(artist = "Kendrick Lamar", album = "DAMN.")

head(DAMN)
```

Bam. Easy peasy. Now you have a sweet data frame ready for a tidy text analysis!


## Multiple albums or songs

Being able to create a dataframe with multiple artists and albums is extremely useful for tidytext analysis. Instead of having to iterate over your data, `add_genius()` is here to assist you. 

Pipe a dataframe with a column for the album artists and album/track information. The argument `type` is used to indicate if the dataframe contains songs or albums

```{r}
# Example with 2 different artists and albums
artist_albums <- tribble(
 ~artist, ~album,
 "J. Cole", "KOD",
 "Sampha", "Process"
)


artist_albums %>%
 add_genius(artist, album)


# Example with 2 different artists and songs
artist_songs <- tribble(
 ~artist, ~track,
 "J. Cole", "Motiv8",
 "Andrew Bird", "Anonanimal"
)

artist_songs %>%
 add_genius(artist, track, type = "lyrics")
```



## Tracklists

I often only know an album name and none of the track titles or I only know the position in the tracklist. For this reason, I created a tool to provide an album tracklist. This function, `genius_tracklist()` takes the arguments `artist` and `album`. Simple enough, right? 

Let's get the tracklist for the original release of **DAMN.**. However, real Kendrick fans know that the album was intended to be listened to in chronological *and* reverse order—as is on the collector's release. 


```{r}
damn_tracks <- genius_tracklist(artist = "Kendrick Lamar", album = "DAMN.")

# Collector's reverse order
damn_tracks %>% 
  arrange(-track_n)
```


